A search method for Gaussian mixtures using MML

### 2017-10-10 articles/gmm-search/ms.tex
Expanding the MML objective function in search paper [wip]

%% This document is part of the snob project
%% Copyright 2017 the authors. All rights reserved.
\documentclass{elsarticle}
\journal{Journal of Artificial Intelligence}
\usepackage{bm,url,amsmath}
%\usepackage{boldsymbol}
\usepackage{verbatim} % for \begin{comment}'ing out sections.

@@

@@

% Have some maaaath.
\def\infinity{\rotatebox{90}{8}}
\def\veccov{\vect{C}}
\def\vecmean{\vect{\mu}}
\def\vectheta{\vect{\theta}}
\def\weight{w}
\def\weights{\vect{\weight}}
\def\datum{y}
\def\data{\vect{\datum}}
\def\likelihood{\mathcal{L}}

@@

\section{MML}
A complex model is not an optimal one, unless its complexity is
justifiable by the added explanatory power. 
There is an established connection between simplicity and truth, as beautifully 
elucidated by the Occam's Razor principle. MML is a statistical technique that formalizes and quantifies this principle.

@@

\section{Message Length}
There are $N$ data points each with $D$ dimensions, which are to be modelled
by $K$ mixture of $D$-dimensional gaussian distributions, each with a relative
weighting $\weight_k$ such that $\sum_{k=1}^{K}\weight_k = 1$. 
The data have the same error value, $y_{err}$, in all $D$ dimensions, for all $N$ observations.

The full expression for the message length is
given by,

@@

\section{The Heuristic}

>>>>

%% This document is part of the snob project
%% Copyright 2017 the authors. All rights reserved.
\documentclass{elsarticle}
\journal{Journal of Artificial Intelligence}
\usepackage{bm,url,amsmath}
%\usepackage{boldsymbol}
\usepackage{color} % for \todo command
\usepackage{verbatim} % for \begin{comment}'ing out sections.

@@

\newcommand{\todo}[1]{\textcolor{red}{#1}}

@@

% Have some maaaath.
\def\infinity{\rotatebox{90}{8}}
\def\veccov{\vect{C}}
\def\vecmean{\vect{\mu}}
\def\vectheta{\vect{\theta}}
\def\weight{w}
\def\weights{\vect{\weight}}
\def\datum{y}
\def\data{\vect{\datum}}
\def\likelihood{\mathcal{L}}
\newcommand{\fisher}[1]{\mathcal{F}\left(#1\right)}
\newcommand{\detfisher}[1]{\left|\fisher{#1}\right|}
\newcommand{\prior}[1]{p\left(#1\right)}

@@

\section{MML}
% \section{Minimum Message Length (MML)}
A complex model is not an optimal one, unless its complexity is
justifiable by the added explanatory power. 
There is an established connection between simplicity and truth, as beautifully 
elucidated by the Occam's Razor principle. MML is a statistical technique that formalizes and quantifies this principle.

@@

\section{Message Length}

The MML principle requires that we encode everything required to reconstruct 
the message, including our prior beliefs on the model parameters. Consider 
that we have $N$ data points each with $D$ dimensions which are to be 
modelled by a finite number of Gaussian mixtures. Therefore the message for 
a finite number of Gaussian mixtures must encode:

\begin{enumerate}
  \item The number of Gaussian mixtures, $K$.
  \item The relative weights $\weights$ of the $K - 1$ Gaussian mixtures. Only
        $K - 1$ weights must be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
  \item The component parameters $\vecmean$ and $\veccov$ for all $K$ Gaussian
        mixtures.
  \item The data, given the model parameters.
  \item The lattice quantisation constant $\kappa(Q)$ for the number of model
        parameters $Q$.
\end{enumerate}

We provide the message length of each component in turn before outlining our
full objective function. We assume a prior on the number of mixtures $K$ of
$\prior{K} \propto 2^{-K}$ \todo{\cite{someone}} such that the length of the 
optimal lossless message to encode $K$ is 
\begin{equation}
	I(K) = -\log{\prior{K}} = K\log{2} + \textrm{constant}.
\end{equation}

We assume a uniform prior on the individual weights $\weight_{k}$. The weights
$\weights$ are drawn from a multinomial distribution such that they can be 
optimally encoded in a message with length\todo{\cite{someone}}:
\begin{equation}
  I(\weights) 
    = \frac{K - 1}{2}\log{N} 
    - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k 
    - \log{\Gamma{\left(K\right)}} \quad .
\end{equation}

In order to properly encode the means $\vecmean$ and covariance matrices
$\veccov$ for all $K$ mixtures, we must encode both our prior belief on 
those parameters and the determinant of the expected Fisher information 
matrix. For the $k$-th mixture the joint message length is
\begin{eqnarray}
  I(\vecmean_k,\veccov_k) &=& -\log{\left(\frac{\prior{{\vecmean_k,\veccov_k}}}{\sqrt{\detfisher{{\vecmean_k,\veccov_k}}}}\right)} \nonumber \\ 
  I(\vecmean_k,\veccov_k) &=& -\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\log{\detfisher{{\vecmean_k,\veccov_k}}}
\end{eqnarray}

\noindent{}and for all mixtures:
\begin{equation}
  I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \quad .
  \label{eq:I_component_params}
\end{equation}

We use an improper uniform prior on the mean $\vecmean$ for all mixtures such
that 
	$\mathcal{U}(\vecmean) = (-\infty, +\infty)$.
However,  the MML principle \todo{demands} proper priors. In practice we make 
this prior proper by assuming the bounds on $\prior{\vecmean}$ are very large, but 
we do so without actually specifying the value of the bounds because they add 
only a constant value to the total message length\todo{\cite{someone}}.
We adopt a conjugate inverted Wishart prior for the covariance matrix of an
individual mixture \cite[e.g., Section 5.2.3. of ][]{Schafer_1997}:
\begin{equation}
  \prior{{\vecmean_k, \veccov_k}} \propto |\veccov_k|^{\frac{D+1}{2}} \quad .
  \label{eq:covariance-prior}
\end{equation}

% Casey: Reword following paragraph because it is clumsy.
Following Eq. \ref{eq:I_component_params}, we require the logarithm of the
determinant of the expected Fisher information matrix for $\vecmean_k$ and
$\veccov_k$ in order to encode the full message length. We approximate the
determinant of the Fisher information $\detfisher{{\vecmean_k, \veccov_k}}$ 
as the product of $\detfisher{\vecmean_k}$ and $\detfisher{\veccov_k}$ 
\cite{Oliver_1996,Roberts_1998}. In order to derive $\detfisher{\vecmean_k}$
we take the second derivative of the log-likelihood function 
$\likelihood(\data|\vecmean,\veccov)$, which yields:
\begin{equation}
  \detfisher{\vecmean_k} = (N\weight_k)^{D}|\veccov_k|^{-1} \quad .
\end{equation}
We make use of an analytical expression for $\detfisher{\veccov_k}$
\cite{Dwyer_1967,Magnus_1988}:
\begin{equation}
  \detfisher{\veccov_k} = (N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \quad .
\end{equation}
Combining these expressions gives our approximation for the determinant of
the expected Fisher information $\detfisher{{\vecmean_k,\veccov_k}}$:
\begin{eqnarray}
  \detfisher{{\vecmean_k,\veccov_k}} & \approx & \detfisher{\vecmean_k} \cdot \detfisher{\veccov_k} \nonumber \\
  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^{D}|\veccov_k|^{-1}(N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
  \label{eq:mixture-component-det-fisher}
\end{eqnarray}

Substituting Eqs. \ref{eq:covariance-prior} and \ref{eq:mixture-component-det-fisher} 
into Eq. \ref{eq:I_component_params} gives the message length to encode the component
parameters $\vecmean$ and $\veccov$ for all $K$ mixtures:

\begin{eqnarray}
  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \nonumber \\
  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{|\veccov_k|^{\frac{D + 1}{2}}}
	+ \frac{1}{2}\sum_{k=1}^{K}\log{\left[(N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right]} \nonumber \\
  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
	-\frac{KD}{2}\log{2} \nonumber \\
  && -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{2D+3}{2}\sum_{k=1}^{K}\log{|\veccov_k|} - \frac{KD}{2}\log{2} 
\end{eqnarray}

If we assume that the data have homoskedastic noise properties, then the 
precision of each measurement $\mathcal{E}$ relates the probability of a
datum $p(\datum_n)$ can be related to the probability density of
the datum, given the model 
$p(\datum_n) = \mathcal{E}^{D}p(y_i|\mathcal{M})$.
In this work we adopt $\mathcal{E} = 0.01$, but we note that the value of
$\mathcal{E}$ has no effect on our inferences.  The message length of encoding
a datum given the model is,
\begin{equation}
  I(\data_n) = -\log{p(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
\end{equation}

\noindent{}and for the entire data:
\begin{eqnarray}
  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .
\end{eqnarray}

The final part of the message is to encode the lattice quantisation constant,
which arises from the approximation to the (so-called) strict MML, where
parameters are quantised into intervals in high dimensional space in order to
be losslessly encoded.  We use an approximation of the lattice constant
\citep[see Sections 5.1.12 and 3.3.4 of ][]{Wallace_2005} such that,
\begin{equation}
  \log\kappa(Q) = \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1 \quad ,
\end{equation}

\noindent{}where $Q$ is the total number of free parameters in the model. Each
Gaussian component in the mixture requires $D$ parameters to specify $\vecmean$,
and a covariance matrix with off-diagonal terms requires $\frac{1}{2}D(D+1)$ 
parameters. Because only $K - 1$ weights need to be encoded, the total number of free
parameters is
\begin{equation}
	Q = K\left[\frac{D(D+3)}{2} + 1\right] - 1 \quad .
	\label{eq:number-of-parameters}
\end{equation}

%  Here

 Each $k$-th mixture is a $D$-dimensional Gaussian that contributes a relative weight $\weight_k$ such that $\sum_{k=1}^{K}\weight_k = 1$. We will assume that the data have homoscedastic noise properties.

The data have the same error value, $y_{err}$, in all $D$ dimensions, for all $N$ observations.

The message length for the data is given by:

Here we introduce the individual components that contribute to the message length.

Combining these expressions, the complete message length for $K$ mixtures is given by

@@

\section{The Heuristic}
% \section{The Heuristic - Method of Moments}
